Demystifying the Computer Science Ph.D. Admission in the US

### 2023-06-10 phd-cs-us.tex
updates

\newtcolorbox{cbox}[1][]
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt,
title=#1 % I would like to make this (one of these in general) assignment optional depending on #1, #2...
}

 \newenvironment{commentbox}[1][]{
 \small
    \begin{cbox}[#1]
 }{
   \end{cbox}
}

@@

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for Vietnamese students interested in CS, it should be applicable to students in various countries interested in STEM.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (see \S\ref{sec:ranking}), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US refers to a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

\begin{commentbox}[]
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few applications in Vietnam (e.g., our CS PhD program has on average 3 applications from Vietnam each year). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{commentbox}

%% Vu: *what's a PhD?*  This [[https://matt.might.net/articles/phd-school-in-pictures/][series of pictures]] from [[https://matt.might.net][Matt Might]] illustrates what a PhD means.
%% \end{tcolorbox}

@@

\begin{commentbox}[]
Vu: We usually decide that a PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you with no funding because you have funding on your own (e.g., you are
supported by your government or have external grants). We justify
our decision with a short summary about your application, where we list
strengths (e.g., well-known undergrad school) and weaknesses (e.g., weak
LORs).
\end{commentbox}

\begin{commentbox}[]
    Hakan: At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as "optional", it is mandatory and they should plan for submitting all those.  
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments/programs do not have the flexibility to waive the application fee, even if they want to.

@@

\begin{commentbox}[]
Vu: Vietnamese students often mention Scopus Q1, which consists of diverse journals from IEEE, Elsevier, and many other publishers unfamiliar to me.  Honestly, I don't know/recognize many of journals listed in Scopus Q1 and. So this might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

\begin{commentbox}[]
Craig: GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{commentbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: It's better to have a good personalized
letter about your own research ability from someone who is less
well-known than a generic/weak letter from a well-known person.

Vietnamese students sometimes go through great length just to get letters from well-known senior professors in their school, but the letters are very generic and weak and therefore carry little value, in fact, red flags. Moreover, a top professor in Vietnamese might not be well-known to US faculty (see more details in \S\ref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors who knows you well and can write a good letter about your research ability.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
\small
Vu: I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Sometime PhD admission committee will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. I have looked at Vietnamese applications (whether they are assigned to me or not) and provide inputs to the reviewers of those applications, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required, though it can help boost your profile
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and wants argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: GMU has Virtual Open House, e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}, which I've co-organized in the last two years. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU CS, students admitted with TA have  4 years of GTA guaranteed and also receive  stipend for the \textbf{first} summer.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: If you're lucky and got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit your application (unless your application has many red flags).
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, Ph.D. applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{tcolorbox}

@@

\begin{enumerate}
    \item   Is an MS degree required for admission to PhD?
    \begin{description}
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{description}

    \item Can I apply to PhD to CS if my undergrad was in Math, Biology, Finance, etc?
    \begin{description}
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned.
    \end{description}

    \item What can you do to increase your admission chance?
    \begin{description}
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

@@

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{tcolorbox}
    \end{description}
\end{enumerate}

>>>>

\newtcolorbox{cbox}[]
{
left=1pt,right=1pt,top=1pt,bottom=1pt,
colback=gray!10,
boxrule=0.5pt
}

 \newenvironment{commentbox}{
 \small
    \begin{cbox}
 }{
   \end{cbox}
}

@@

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for Vietnamese students interested in CS, it should be applicable to students in various countries interested in STEM.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (see \S\ref{sec:ranking}), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

\begin{commentbox}
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few applications in Vietnam (e.g., our CS PhD program has on average 3 applications from Vietnam each year). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{commentbox}

%% Vu: *what's a PhD?*  This [[https://matt.might.net/articles/phd-school-in-pictures/][series of pictures]] from [[https://matt.might.net][Matt Might]] illustrates what a PhD means.
%% \end{commentbox}

@@

\begin{commentbox}
Vu: We usually decide that a PhD candidate is either (i) admit with funding (TA or RA) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we don't. In some rare cases, we may admit
you with no funding because you have funding on your own (e.g., you are
supported by your government or have external grants). We justify
our decision with a short summary about your application, where we list
strengths (e.g., well-known undergrad school) and weaknesses (e.g., weak
LORs).
\end{commentbox}

\begin{commentbox}
    Hakan: At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as "optional", it is mandatory and they should plan for submitting all those.  
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments/programs do not have the flexibility to waive the application fee, even if they want to.

@@

\begin{commentbox}
Vu: Vietnamese students often mention Scopus Q1, which consists of diverse journals from IEEE, Elsevier, and many other publishers unfamiliar to me.  Honestly, I don't know/recognize many of journals listed in Scopus Q1 and. So this might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

\begin{commentbox}
Craig: GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{commentbox}

@@

\begin{commentbox}
Vu: It's better to have a good personalized
letter about your own research ability from someone who is less
well-known than a generic/weak letter from a well-known person.

Vietnamese students sometimes go through great length just to get letters from well-known senior professors in their school, but the letters are very generic and weak and therefore carry little value, in fact, red flags. Moreover, a top professor in Vietnamese might not be well-known to US faculty (see more details in \S\ref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors who knows you well and can write a good letter about your research ability.
\end{commentbox}

@@

\begin{commentbox}
\small
Vu: I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\end{commentbox}

@@

\begin{commentbox}
Vu: Sometime PhD admission committee will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. I have looked at Vietnamese applications (whether they are assigned to me or not) and provide inputs to the reviewers of those applications, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}

@@

\begin{commentbox}
Vu: The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required, though it can help boost your profile
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{commentbox}

@@

\begin{commentbox}
Vu: At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and wants argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{commentbox}

@@

\begin{commentbox}
Vu: GMU has Virtual Open House, e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}, which I've co-organized in the last two years. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{commentbox}

@@

\begin{commentbox}
Vu: TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\end{commentbox}

@@

\begin{commentbox}
At GMU CS, students admitted with TA have  4 years of GTA guaranteed and also receive  stipend for the \textbf{first} summer.
\end{commentbox}

@@

\begin{commentbox}
Vu: If you're lucky and got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit your application (unless your application has many red flags).
\end{commentbox}

@@

\begin{commentbox}
At GMU, Ph.D. applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{commentbox}

@@

\begin{enumerate}   
    \item What can you do to increase your admission chance?
    \begin{description}
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

@@

    \item Can I apply to PhD to CS if my undergrad was not in CS or any related areas?
    \begin{description}
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able leverage this to make your profile stand out.
    \end{description}

    \item   Is an MS degree required for admission to PhD?
    \begin{description}
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{description}

@@

\begin{commentbox}
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{commentbox}
    \end{description}
\end{enumerate}
